---
title: "*dtree* vignette"
author: "P.J. Dodd"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache=FALSE)
```

#Note

This vignette is a work in progress...

TODO:
- currently the dot strings need search/replace "" to '' to be valid DOT having written out with `cat()`

#Introduction

- decision trees in HEE & other packages
- Rationale:
- free, simple
- arbitrarily correlated variables resulting from outputs from MCMC analyses
- patient-level & statistical models
- onward uncertainty VoI analysis


#Installation

Install `devtools` by typing `install.packages('devtools')` if needed.

Then type

```{r,eval=FALSE}
devtools::install_github('petedodd/dtree',dependencies=TRUE,build_vignettes=TRUE)
```


#Use

```{r,eval=TRUE}
library(dtree)
```

A decision tree is defined with the following syntax:
```{r,eval=TRUE}
## example syntax
trss0 <- "
\\nodes...(this line is  comment)
\\ (this line is another comment) syntax:
\\ nodename ['label | cost variable | QoL variable']
a[ 'description a | costA | qolA'] \\ a node
b[ 'description b | costB | qolB']
c[  'description c | costC | qolC']
d[  'description d | costD | qolD']
e[  'description e | costE | qolE']
f[  'description f | costF | qolF']
g[  'description g | costG | qolG']
h[  'description h | costH | qolH']
i[  'description i | costI | qolI']
\\edges
\\ from_nodename -> to_nodename['probability variable for this edge']
a -> b ['p1']
a -> c ['1-p1']
c -> e ['p2']
c -> f['p3']
c -> g['p4']
b -> h['p5']
b -> i['p6']
b -> d['p7']
\\ (whitespace is ignored in parsing for calculation)
"
```

To view the tree and check it is as intended:
```{r}
dtree(trss0,view=TRUE) ## visualize the overall tree
```

To parse the tree and generate functions for analysis:
```{r}
decT <- dtree(trss0,view=FALSE) ##analyse tree
str(decT)
```
One can see that `decT` is returned as a list with the following quantities:

- `coststring`: a string with mean cost of the tree in terms of its variables
- `qolstring`: a string with mean QoL of the tree in terms of its variables
- `costfun`: a function generating the mean cost of the tree applied to PSA data
- `qolfun`: a function generating the mean QoL of the tree applied to PSA data
- `root`: the name of the root node of the tree
- `dot`: a text string specifying the graph visualized by `dtree(trss0,view=TRUE)` in the DOT languange. This can be exported for generation of pdf or other output via GraphViz.
- `intputstring`: the string used as input

Example data is specified as a list:

```{r}
costpars <- list(costA=1,costB=2,costC=3,costD=1,costE=2,costF=3,costG=2,costH=3,costI=3,
                 qolA=.4,qolB=1,qolC=0.9,qolD=.4,qolE=1,qolF=0.9,qolG=.4,qolH=1,qolI=0.9,
                 p1=1/3,p2=1/3,p3=1/3, p4=1/3,p5=1/3,p6=1/3, p7=1/3) #a list of input parameters

decT$costfun(costpars) #calculate the mean cost for these parameters
decT$qolfun(costpars) #calculate the mean QoL for these parameters
```
N.B. arithmetic for probabilities etc can be done in the specification of the variables of the tree: calculations are managed via strings and probabilities are wrapped in parentheses.  E.g. `1-p1` as above is fine as long as an input variable `p1` is supplied, and is wrapped as `(1-p1)` in the strings.

N.B. there is no checking that probabilities sum to 1.

However, these functions can also be applied to `data.frame`s of PSA data:
```{r}
costparz <- as.data.frame(costpars) #list to data.frame
costparz <- costparz[rep(1,1e2),] #a data.frame with 100 identical rows
costparz$cost <- decT$costfun(costparz) # output a vector of length 100 & augment input
costparz$qol <- decT$qolfun(costparz) #augment input data with outputs
summary(costparz$cost) #check makes sense
```

#Less trivial examples

##Correlated variables

Let us take the above example but include some uncertainty around costs:

```{r}
csts <- paste0('cost',LETTERS[1:9]) # cost variables
for(cst in csts)
  costparz[,cst] <- costparz[,cst] * (1+0.2*runif(nrow(costparz))) #perturb
costparz$costB <- costparz$costB + costparz$costC # make sure these are correlated
cor(costparz$costB,costparz$costC) #correlation
plot(data=costparz,costB ~ costC) #view
qls <- paste0('qol',LETTERS[1:9]) # QoL variables
for(ql in qls)
  costparz[,ql] <- costparz[,ql] * runif(nrow(costparz)) #perturb
# renew cost data
costparz$cost <- decT$costfun(costparz) # costs
costparz$qol <- decT$qolfun(costparz) # QoL
plot(data=costparz,cost ~ qol,asp=1,xlim=c(0,2)) #view
```

todo:

- simple
- use lhs
- do some plotting (need to write functions)

##Individual-level simulation

todo:

- simple example
- patient-level covariates
- statistical model for probabilities
